---
title: "IPE Paper: Kazakhstan and Foreign Direct Investment"
author: "Daniel Shapiro"
date: "PSCI 6402"
output: pdf_document
header-includes:
  - \usepackage{setspace}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(readr)
library(readxl)
```

```{r}
industries <- read.csv("wiiw-fdi-database.csv")
partners <- read.csv("wiiw_partners.csv")
```

```{r}
editeddata <- industries %>%
  filter(Unit == "EUR m")
```

```{r}
editedpartners <- partners %>% 
  filter(Unit == "EUR m") %>%
  filter(Partner %in% c("China", "Russia", "United States")) %>%
  select(-c(Tree.Level.1, Country.Code, Country, Partner.Code, Unit, Footnotes, Sources)) %>%
  pivot_longer(!c(Direction, Partner), names_to = "Year", values_to = "Value")

editedpartners$Year <- gsub("X_", "", editedpartners$Year)
editedpartners$Year <- as.numeric(editedpartners$Year)
editedpartners$Value <- as.numeric(editedpartners$Value)
```

```{r stock}
fdistock <- editedpartners %>%
  filter(Direction == "FDI inward stock")
fdiinflow <- editedpartners %>%
  filter(Direction == "FDI inflow")
```

\doublespacing

Kazakhstan, the largest country in Central Asia and the 9th-largest in the world, has a unique geopolitical and geoeconomic profile. A former Soviet republic with a large number of ethnic Russian citizens and a relatively authoritarian government, Kazakhstan has enjoyed close ties with Russia since the dissolution of the Soviet Union in 1991, joining the Russia-led Eurasian Economic Union (EAEU) and the Collective Security Treaty Organization (CSTO), as well as the Shanghai Cooperation Organization (SCO), among other organizations. Additionally, given its 953-mile long border with China, Kazakhstan has steadily developed its political and economic ties with its large eastern neighbor over the past 30 years. Trade between the two countries has increased rapidly, and China's Belt and Road Initiative (BRI) features Kazakhstan as a key transit country for goods travelling between China and Europe. Given its positioning between Russia to its north and China to its east, it is no surprise that Kazakhstan has become a key point of focus for both of these powerful states.

However, looking at one measure of cooperation -- foreign direct investment (FDI) -- an interesting trend emerges. While China and Russia do work closely with Kazakhstan in many areas, the United States far outstrips these two powers in FDI stock in Kazakhstan, especially beginning in 2014 -- arguably the beginning of the current era of overt great power competition. Figure 1 shows this trend:

*Figure 1:*
  
```{r}
ggplot(fdistock, aes(x = Year, y = Value, color = Partner)) +
  geom_point() +
  geom_line() +
  labs(title = "Kazakhstan FDI Stock, 2003-2021",
       subtitle = "Data from wiiw",
       y = "FDI Stock in EUR m",
       x = "Year")
```

What explains this rise in American investment and the gulf between America's presence and China's and Russia's? There are certainly economic factors at work here -- the 2014 oil price crash and ensuing depreciation of the tenge likely influenced this phenomenon, for example. However, there are political explanations as well. I identify several hypotheses related to different potential political factors that could impact the increased gulf between American FDI stock and Chinese and Russian FDI stock in Kazakhstan, with some centered around internal Kazakh politics, others on American politics, and others on global politics. Through process tracing, we can see that aspects of all three were at work, but ________.

### Hypotheses

This paper takes three hypotheses on the political reasons for heightened US FDI stock in Kazakhstan (relative to Chinese and Russian) and analyzes them through process tracing to determine their veracity. I find that _____

* H1: **Internal: Kazakhstan.** Political changes inside Kazakhstan led to increased American investment in comparison to Chinese and Russian investment.
* H2: **Internal: USA.** Political changes inside the US led to increased American investment in comparison to Chinese and Russian investment.
* H3: **External: International.** Political changes in the international arena led to increased American investment in comparison to Chinese and Russian investmentin Kazakhstan.

### Background and Literature Review

This paper is situated within a number of important debates in political science, the most relevant of which deal with dynamics related to foreign direct investment. First, there is the question of why companies choose to engage in foreign direct investment in general. Two famous models that offer explanations of this phenomenon are Dunning's OLI (Ownership, Location, Internalization) paradigm[^1] and Markusen's Knowledge Capital model[^2]. While both of these models offer persuasive evidence as to why firms *in general* choose to pursue foreign direct investment, they are less well-suited to answering the political side of the FDI issue and certainly do not address differentiation in approaches to FDI in a given country over time.

On a more political level, we know a number of trends that help drive the conversation on FDI. For one, on a rudimentary level, we know that the vast majority of FDI outflow comes from countries in the Global North, thus making the US's high level of FDI in Kazakhstan not entirely surprising. Despite a dip during the COVID-19 pandemic, over the past decade, "developed economies" have accounted for anywhere between 60 and 80 percent of global outflows, with most yearly figures around 75 percent.[^3] This trend has been the case for some time; Walter and Sen (2009), for example, report that developed economies made up 85 percent of global FDI outflow in 2005. Given that UNCTAD does not consider China a "developed economy," there is some logic behind China's relatively low position in comparison to that of the United States. Russia, however, is considered "developed" by UNCTAD -- but despite its proximity to Kazakhstan, it still has relatively low levels of FDI stock in the country. Overall, while this basic explanation helps frame some global trends, it is far from sufficient as an explanatory variable as to why American FDI stock in Kazakhstan grew so substantially over the past few years in comparison to China and Russia.

Moving to more substantive analysis, many scholars have looked at changing internal aspects of countries and their effects on trends in FDI. One key debate in the FDI literature is over whether or not more democratic or more authoritarian countries attract more FDI. Some scholarship indicated that authoritarian countries were more attractive targets for multi-national corporations (MNCs); authors such as Putnam (1988), for example, posit that democracies' "two-level game," in which leaders have to simultaneously play an international game and a domestic political game due to their accountability to their electorate when negotiating, can potentially present too many hurdles and too many interested parties with potential veto power for interested MNCs. On the other hand, Jensen (2003) argues the opposite, using empirical evidence to show that democratic institutions lower political risk and make investors more likely to want to invest.[^4] Li (2009) also shows that democracies are less likely to turn to asset expropriation than autocracies.  These theories likely have some degree of truth; MNCs may like the lack of bureaucratic and citizen oversight that authoritarian governments offer but dislike the arbitrariness of the regime and the risk involved with subpar protection of property rights. Given that Kazakhstan is unequivocally authoritarian -- and especially so during the initial period in which we see divergence increasing -- this general debate will shape this paper as well.

On a more granular level, specific attributes of a state can increase or decrease its attractiveness to foreign investors. This area of research relates directly to H1 -- to what degree do changing internal dynamics in Kazakhstan determine the outcome? There are several areas that scholars posit may impact MNCs' likelihoods to invest in a given location. High human capital is one area (Mankiw et. al 1992); Dunning (1993) also refers to geographic position, regional trade liberalization and physical infrastructure as key factors. Tax policy is another important aspect; hile some argue that tax incentives have a heavy impact on MNCs' likelihoods to invest in one place over another, others (Kobrin 1987; Morisset and Pirnia 2000; Walter and Sen 2009) take more measured approaches. Low labor costs could act as another factor that could drive an MNC to invest in a given state; weak environmental regulation could also have an impact. As briefly alluded to previously, institutions also help -- there is a large amount of scholarship on the idea that solid courts and institutions that can protect companies' rights help lead to increased investment (Levy and Spiller 1994; Li and Resnick 2003; Jensen 2006; Staats and Biglaiser 2012). Finally, for resource-rich countries like Kazakhstan, new resource discoveries or increased allowances for foreign companies to access said resource deposits could have a major impact as well. Based on conclusions from this literature, we will be able to draw conclusions as to the veracity of H1.

Additionally, internal political dynamics in the host country can have an impact on FDI. Beazer and Blake (2018) utilize a number of different data sets and find that "host institutions' attractiveness depends on firms' home environment," as "home country institutions shape firms' practices and capabilities, thus helping to determine the environments that firms are best prepared to face abroad." Thus, it is possible that the political atmosphere or institutions changed in the United States during this period, causing firms to change their course. If H2 is correct, however, it would more likely be due to shorter-term factors than longer-term ones: rather than wholesale institutional change, the US government may have instituted policies that changed how firms interact with countries such as Kazakhstan, leading to a rise in American investment. 

Finally, as posited in H3, the international political environment could have an impact on this recent pattern in these countries' respective FDI stocks in Kazakhstan. Delios and Henisz (2003) write that MNCs' international experience in general can impact their strategies. Given the tumult in the region at the time -- especially with regard to the Russian annexation of Crimea, the beginning of the war in Ukraine and ensuing sanctions policies enacted against Russia (Kazakhstan's neighbor) by many governments around the world -- it is possible that US-based MNCs reacted to external stimuli differently than Russia-based ones or China-based ones. While American firms may have decided that further investment in Kazakhstan was smart, Chinese or Russian ones may not have.

Using this background, this paper makes a couple of key contributions. For one, it uses a single-country case study to elucidate detailed procedures and policies surrounding reasons behind changes in countries' FDI positions. Additionally, it applies FDI literature to a heavily contested part of the world (Central Asia) and utilizes process tracing to uncover key drivers of trends in FDI. Especially in the current climate of great power competition (and almost outright conflict, in the case of the US and Russia), understanding the political factors behind trends in FDI in a country coveted in different ways by three major powers -- the US, Russia, and China -- can help shed more light not only on the FDI literature as a whole, but also on how states make decisions regarding foreign investment, how authoritarian states and democracies work with MNCs, and how the investment climate reacts to change. From a theoretical standpoint, the lessons learned in this paper are useful of course for scholars interested in FDI and also for those interested in Kazakhstan and Central Asia, international relations, and smaller state strategies in periods of great power competition.

### Data and Methods

This paper relies on data from The Vienna Institute for International Economic Studies (wiiw), an Austria-based organization that provides, among other things, the wiiw FDI database. The database markets itself as a "comprehensive and unique database on Foreign Direct Investment activities related to 23 countries in Central, East and Southeast Europe." All data on FDI in Kazakhstan is taken from this database.

Methodologically, this study uses process tracing to analyze dynamics related to foreign direct investment in a single-country case study (Kazakhstan). Lynch (2005) provides a good defense for the efficacy of single-country case studies, writing that "within-country comparisons make it possible to test hypotheses with a high degree of internal validity, since they essentially control for possibly confounding covariates at the level of national institutions, policy-making styles, resource endowments, positioning in the world economy, cultural beliefs, or the like." According to scholars such as Lynch, a more complex notion of causality "...demands that good theories not only 'predict' the apposite political outcome, but also get the causal process that generated it right." Bennett and George (1993) write that single case studies can be used as "exemplars" for exploratory research and hypothesis-generating research. Lynch (2005) also argues that single-country case studies can be used for "tough cases" which pose a particular challenge to existing theoretical frameworks and models.



```{r}
ggplot(fdiinflow, aes(x = Year, y = Value, color = Partner)) +
  geom_point() +
  geom_line() +
  labs(title = "Kazakhstan FDI Inflow, 2003-2021",
       subtitle = "Data from wiiw",
       y = "FDI Inflow in EUR m",
       x = "Year")
```


[^1]: John Dunning, Explaining International Production (London: Unwin Hyman, 1988).
[^2]: James R. Markusen, Multinational Firms and the Theory of International Trade (Cambridge:
MIT Press, 2002).
[^3]: Data from UNCTAD. Link: https://hbs.unctad.org/foreign-direct-investment/#:~:text=The%20share%20of%20developed%20economies,cent%20to%2025.7%20per%20cent. 
[^4]: See also Li, 2006; Li & Resnick, 2003. 