---
title: "IPE Paper: Kazakhstan and Foreign Direct Investment"
author: "Daniel Shapiro, PSCI 6402"
output: pdf_document
header-includes:
  - \usepackage{setspace}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(readr)
library(readxl)
```

```{r}
industries <- read.csv("industrial.csv")
partners <- read.csv("wiiw_partners.csv")
```

```{r}
editeddata <- industries %>%
  filter(Unit == "EUR m")
```

```{r}
editedpartners <- partners %>% 
  filter(Unit == "EUR m") %>%
  filter(Partner %in% c("China", "Russia", "United States")) %>%
  select(-c(Tree.Level.1, Country.Code, Country, Partner.Code, Unit, Footnotes, Sources)) %>%
  pivot_longer(!c(Direction, Partner), names_to = "Year", values_to = "Value")

editedpartners$Year <- gsub("X_", "", editedpartners$Year)
editedpartners$Year <- as.numeric(editedpartners$Year)
editedpartners$Value <- as.numeric(editedpartners$Value)
```

```{r stock}
fdistock <- editedpartners %>%
  filter(Direction == "FDI inward stock")
fdiinflow <- editedpartners %>%
  filter(Direction == "FDI inflow")
```

\doublespacing

Kazakhstan, the largest country in Central Asia and the 9th-largest in the world, has a unique geopolitical and geoeconomic profile. A former Soviet republic with a large number of ethnic Russian citizens and a relatively authoritarian government, Kazakhstan has enjoyed close ties with Russia since the dissolution of the Soviet Union in 1991, joining the Russia-led Eurasian Economic Union (EAEU) and the Collective Security Treaty Organization (CSTO), as well as the Shanghai Cooperation Organization (SCO), among other organizations. Additionally, given its 953-mile long border with China, Kazakhstan has steadily developed its political and economic ties with its large eastern neighbor over the past 30 years. Trade between the two countries has increased rapidly, and China's Belt and Road Initiative (BRI) features Kazakhstan as a key transit country for goods travelling between China and Europe. Given its positioning between Russia to its north and China to its east, it is no surprise that Kazakhstan has become a key area for both states. However, looking at one measure of cooperation -- foreign direct investment (FDI) -- an interesting trend emerges. While China and Russia do work closely with Kazakhstan in many areas, the United States far outstrips these two powers in FDI stock in Kazakhstan, especially beginning in 2014 -- arguably the beginning of the current era of overt great power competition. Figure 1 shows this trend:

*Figure 1:*
  
```{r, out.width = '65%', fig.align = 'left'}
ggplot(fdistock, aes(x = Year, y = Value, color = Partner)) +
  geom_point() +
  geom_line() +
  labs(title = "Kazakhstan FDI Stock, 2003-2021",
       subtitle = "Data from wiiw",
       y = "FDI Stock in EUR m",
       x = "Year") +
  theme(plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))
```

What explains this rise in American investment and the gulf between America's presence and China's and Russia's? There are certainly economic factors at work here -- the 2014 oil price crash and ensuing depreciation of the tenge likely influenced this phenomenon, for example. There are political explanations as well; I identify several hypotheses related to different potential political factors that could impact the increased gulf between American FDI stock and Chinese and Russian FDI stock in Kazakhstan, with some centered around internal Kazakhstani politics, others on American politics, and others on global politics. The most important hypothesis, however, is that the onset of the Russo-Ukrainian War in February 2014 spurred increased American MNC investment in Kazakhstan. Through process tracing, we can see that this factor indeed was at work and we uncover some underlying mechanisms as to why and how this is the case. However, process tracing also shows that aspects of all aforementioned hypotheses were at work.

### Hypotheses

This paper takes three hypotheses on the political reasons for heightened US FDI stock in Kazakhstan (relative to Chinese and Russian) and analyzes them through process tracing to determine their veracity. This paper predicts, as stated by the main hypothesis (H1), that global phenomena -- most notably, the onset of the Russo-Ukrainian War in February 2014 -- led many American MNCs to increase their investments in Kazakhstan in comparison to Russian and Chinese MNCs. However, we test whether concerted efforts by the Nazarbayev regime to attract additional outside investment in Kazakhstan's oil and gas industry and US-based political decisions played some role as well.

* H1: **External: International.** Political changes in the international arena led to increased American investment in comparison to Chinese and Russian investment in Kazakhstan. These changes could be tied either a) to the increased level of political risk in neighboring Russia, causing American companies to move south across the border, or b) to political shifts surrounding American investments in other large hydrocarbon-producing countries, leading American MNCs to invest more heavily in Kazakhstan. Most notably, we look at the *breakout of war* between Russia and Ukraine in February 2014 as a motivating factor.
* H2: **Internal: Kazakhstan.** Political changes inside Kazakhstan led to increased American investment in comparison to Chinese and Russian investment. These changes could include a) liberalization in the Kazakhstani regime, both initially during the Nazarbayev era, and in the present day under Kassym-Jomart Tokayev; b) legislation passed that increases outside access to Kazakhstan's mineral wealth, or c) tightened controls on corruption.
* H3: **Internal: USA.** Political changes inside the US led to increased American investment in comparison to Chinese and Russian investment. These could be, for example, that the US government became more interested in encouraging foreign direct investment in general beginning in 2014, especially in "battleground" areas as American relations with other great powers grew more antagonistic. 

These potential factors are certainly not mutually exclusive, and multiple factors can play important roles simultaneously. Below, I provide some background for the topic, the aforementioned hypotheses, and for the scholarly discourse in which this paper finds itself.

### Background and Literature Review

This paper is situated within a number of important debates in political science, the most relevant of which deal with dynamics related to foreign direct investment. First, there is the question of why companies choose to engage in foreign direct investment in general. Two famous models that offer explanations of this phenomenon are Dunning's OLI (Ownership, Location, Internalization) paradigm[^1] and Markusen's Knowledge Capital model[^2]. While both of these models offer persuasive evidence as to why firms *in general* choose to pursue foreign direct investment, they are less well-suited to answering the political side of the FDI issue and certainly do not address differentiation in approaches to FDI in a given country over time, or how FDI trends change in times of conflict.

On a more political level, we know a number of trends that help drive the conversation on FDI. For one, on a rudimentary level, we know that the vast majority of FDI outflow comes from countries in the Global North, thus making the US's high level of FDI in Kazakhstan not entirely surprising. Despite a dip during the COVID-19 pandemic, over the past decade, "developed economies" have accounted for anywhere between 60 and 80 percent of global outflows, with most yearly figures around 75 percent.[^3] This trend has been the case for some time; Walter and Sen (2009), for example, report that developed economies made up 85 percent of global FDI outflow in 2005. Given that UNCTAD does not consider China a "developed economy," there is some logic behind China's relatively low position in comparison to that of the United States. Russia, however, is considered "developed" by UNCTAD -- but despite its proximity to Kazakhstan, it still has relatively low levels of FDI stock in the country. Overall, while this basic explanation helps frame some global trends, it is far from sufficient as an explanatory variable as to why American FDI stock in Kazakhstan grew so substantially beginning in 2014 in comparison to China and Russia.

Moving to more substantive analysis, many scholars have looked at changing internal aspects of countries and their effects on trends in FDI. One key debate in the FDI literature is over whether or not more democratic or more authoritarian countries attract more FDI. Some scholarship indicated that authoritarian countries were more attractive targets for multi-national corporations (MNCs); authors such as Putnam (1988), for example, posit that democracies' "two-level game," in which leaders have to simultaneously play an international game and a domestic political game due to their accountability to their electorate when negotiating, can potentially present too many hurdles and too many interested parties with potential veto power for interested MNCs. On the other hand, Jensen (2003) argues the opposite, using empirical evidence to show that democratic institutions lower political risk and make investors more likely to want to invest.[^4] Li (2009) also shows that democracies are less likely to turn to asset expropriation than autocracies.  These theories likely have some degree of truth; MNCs may like the lack of bureaucratic and citizen oversight that authoritarian governments offer but dislike the arbitrariness of the regime and the risk involved with subpar protection of property rights. Given that Kazakhstan is unequivocally authoritarian -- and especially so during the initial period in which we see divergence increasing -- this general debate will shape this paper as well.

On a more granular level, specific attributes of a state can increase or decrease its attractiveness to foreign investors. This area of research relates directly to H2 -- to what degree do changing internal dynamics in Kazakhstan determine the outcome? There are several areas that scholars posit may impact MNCs' likelihoods to invest in a given location. High human capital is one area (Mankiw et. al 1992); Dunning (1993) also refers to geographic position, regional trade liberalization and physical infrastructure as key factors. Tax policy is another important aspect; hile some argue that tax incentives have a heavy impact on MNCs' likelihoods to invest in one place over another, others (Kobrin 1987; Morisset and Pirnia 2000; Walter and Sen 2009) take more measured approaches. Low labor costs could act as another factor that could drive an MNC to invest in a given state; weak environmental regulation could also have an impact. As briefly alluded to previously, institutions also help -- there is a large amount of scholarship on the idea that solid courts and institutions that can protect companies' rights help lead to increased investment (Levy and Spiller 1994; Li and Resnick 2003; Jensen 2006; Staats and Biglaiser 2012). Finally, for resource-rich countries like Kazakhstan, new resource discoveries or increased allowances for foreign companies to access said resource deposits could have a major impact as well. Based on conclusions from this literature, we will be able to draw conclusions as to the veracity of H1.

Additionally, internal political dynamics in the home country can have an impact on FDI. Beazer and Blake (2018) utilize a number of different data sets and find that "host institutions' attractiveness depends on firms' home environment," as "home country institutions shape firms' practices and capabilities, thus helping to determine the environments that firms are best prepared to face abroad." Li (2008) discusses this factor in the context of war, writing that "home governments are more likely to encourage FDI to go to their military allies."[^5] Thus, it is possible that the political atmosphere or institutions changed in the United States during this period, causing firms to change their course. If H3 is correct, however, it would more likely be due to shorter-term factors than longer-term ones: rather than wholesale institutional change, the US government may have instituted policies that changed how firms interact with countries such as Kazakhstan, leading to a rise in American investment. 

Finally, as posited in H1, the international political environment could have an impact on this recent pattern in these countries' respective FDI stocks in Kazakhstan. Delios and Henisz (2003) write that MNCs' international experience in general can impact their strategies, while Nigh (1985) writes about international conflict and its impact on US MNC behavior during the Cold War period, finding that international and intranational conflicts reduce US investment. Given the tumult in the region at the time -- especially with regard to the Russian annexation of Crimea, the beginning of the war in Ukraine and ensuing sanctions policies enacted against Russia (Kazakhstan's neighbor) by many governments around the world -- it is possible that US-based MNCs reacted to external stimuli differently than Russia-based ones or China-based ones. Maybe American companies worried about getting forced out of Russia due to sanctions decided to uproot themselves and move quickly south across the border into Kazakhstan. Or, perhaps, while American firms may have simply decided that given the climate at the time, further investment in Kazakhstan was smart, Chinese or Russian ones may not have made the same decision.

Using this background, this paper makes a couple of key contributions. For one, it uses a single-country case study to elucidate detailed procedures and policies surrounding reasons behind changes in countries' FDI positions, especially during periods of war. The relationship between FDI and war is understudied in the IPE literature: in general, there is the understanding that increased political risk leads to decreased foreign direct investment, but Li (2008) writes that it is "...imperative that empirical inquiry be expanded."[^6] Bussmann (2010) formulates some basic statements on how militarized conflicts inhibit foreign investment, while some authors, like Johns and Wellhausen (2016), take an approach based on property rights protection and expropriation. They write that MNCs that are more tightly tied in to other firms in the host country are less likely to suffer property expropriation than those which are less tightly tied to other firms. However, the mechanisms behind firms' specific actions during periods of conflict are not covered much at all in the literature, and while Li's statement came in 2008, there is not as much new literature on the topic as one would perhaps expect. Thus, this paper contributes a great deal in this area.

Additionally, this paper applies the FDI literature that does exist to a heavily contested part of the world (Central Asia) and utilizes process tracing to uncover key drivers of trends in FDI. Especially in the current climate of great power competition, regardless of specific military situations, understanding the political factors behind trends in FDI in a country coveted in different ways by three major powers -- the US, Russia, and China -- can help shed more light not only on the FDI literature as a whole, but also on how states make decisions regarding foreign investment, how authoritarian states and democracies work with MNCs, and how the investment climate reacts to change. From a theoretical standpoint, the lessons learned in this paper are useful of course for scholars interested in FDI and also for those interested in dynamics of war, Kazakhstan and Central Asia, international relations, and smaller state strategies in periods of great power competition.

### Data and Methods

This paper relies on data from The Vienna Institute for International Economic Studies (wiiw), an Austria-based organization that provides, among other things, the wiiw FDI database. The database markets itself as a "comprehensive and unique database on Foreign Direct Investment activities related to 23 countries in Central, East and Southeast Europe." All data on FDI in Kazakhstan is taken from this database. This study largely focuses on FDI Stock as an indicator, rather than FDI flow; for arguments as to why FDI stock is a superior indicator to FDI flow, see Kerner (2014).

Methodologically, this study uses process tracing to analyze dynamics related to foreign direct investment in a single-country case study (Kazakhstan). Process tracing can be defined as a key qualitative method that “examines diagnostic pieces of evidence … with the goal of achieving and refining causal inference.”[^7] In this case, potential “diagnostic pieces of evidence” are numerous: many different factors can influence firms' decisions to invest in one country or another, and determining which ones a) actually play a role in the outcome and b) are relevant to this paper requires discretion. Thus, this paper analyzes one main hypothesis and two alternative hypotheses that could impact the observed outcome. This study by no means excludes the possibility that other factors influenced the outcome; however, for purposes of this paper, variables are limited to those directly relating to the aforementioned hypotheses.

In this research, I utilize several data sources, including news articles, opinion pieces, and academic commentaries, as well as journalistic and academic interviews. Additionally, I utilize official statements, meeting summaries, press briefings, and other official correspondence related to FDI. Given that Kazakhstan is a relatively authoritarian and closed country, in-depth primary source-based analysis of the political system is relatively difficult to find, but for this paper, available data is sufficient to gain important insights. Through process tracing, I show how my work supports, partly supports, or does not support my hypotheses. 

Why a single-country case study? Lynch (2005) provides a good defense for the efficacy of single-country case studies, writing that "within-country comparisons make it possible to test hypotheses with a high degree of internal validity, since they essentially control for possibly confounding covariates at the level of national institutions, policy-making styles, resource endowments, positioning in the world economy, cultural beliefs, or the like." According to scholars such as Lynch, a more complex notion of causality "...demands that good theories not only 'predict' the apposite political outcome, but also get the causal process that generated it right." Bennett and George (1997) write that single case studies can be used as "exemplars" for exploratory research and hypothesis-generating research. Lynch (2005) also argues that single-country case studies can be used for "tough cases" which pose a particular challenge to existing theoretical frameworks and models. Kazakhstan certainly presents a unique case. Located between Russia and China but with definite ties to the West and bountiful natural resource deposits, Kazakhstan deserves individual analysis. Additionally, given that this paper utilizes process tracing to make conclusions about trends in FDI sources, keeping the paper limited to only Kazakhstan allows for deep description of causal-process mechanisms; bringing in other countries would likely dilute more than help the argument. 

### Predicted Empirical Evidence and Implications

This paper mostly looks at political factors that could impact these trends in American, Chinese, and Russian FDI Stock in Kazakhstan. There are also undoubtedly economic factors at play as well: Given that Kazakhstan's economy is heavily dependent on natural resource rents, one would suspect that shifts in oil prices and the resulting effects on Kazakhstan's currency and economy will have some effect on FDI trends. However, given that this paper focuses on conflict and other political factors, this section will discuss potential empirical evidence that could help support or not support various hypotheses.

Starting with H1, there are many bits of evidence that could indicate H1's veracity. Most obviously, there could be "smoking gun" evidence: for example, in the most recent round of conflict between Russia and Ukraine that began in February 2022, it was openly reported that many MNCs were moving their operations from Russia to Kazakhstan.[^8] If we find out from research that American MNCs moved en masse from Russia to Kazakhstan in 2014 as well, this could be a major piece of supporting evidence.

However, further research would still be required in order to understand the true depth of such moves, and how much of an impact this cross-border migration would have on FDI stock figures. The FDI literature takes as a given that FDI does not move quickly, and moving billions of dollars worth of FDI stock across an international border may not take place over a short period of time. While some companies certainly may have moved due to conflict, without further research, we cannot be certain a) the degree to which these companies actually comprise a meaningful percentage of Kazakhstan's US-based FDI stock and b) how long it took for them to relocate their operations. Additionally, as shown below by Figure 2, most of Kazakhstan's FDI stock since 2010 has been in "mining and R&D," which, according to wiiw, is largely focused on various steps in the labor-intensive petroleum extraction process.

*Figure 2*:[^9]


```{r}
industry_stock <- industries %>%
  filter(Direction == "FDI inward stock") %>%
  filter(Unit %in% c("in % of total", "in % of total manufacturing")) %>%
  select(-Unit, -Sources, -Tree.Level.1, -Country.Code, -Country, -Direction, -Classification, -Digit, -Footnotes, -X_2000, -X_2001, -X_2002, -X_2003, -X_2004, -X_2005, -X_2006, -X_2007, -X_2008, -X_2009)
# Took out missing data, pivoted, and am now removing excess "manufacturing" breakdown

industry_stock <- industry_stock[2:17, ]

# Now I'm going to combine B Mining and Quarrying with M Professional, scientific and 
#technical activities. There was a reclassification in 2016. First, though, I have to pivot.

industry_stock <- industry_stock %>%
  pivot_longer(!Industry.Product, names_to = "Year", values_to = "Value") %>%
  pivot_wider(names_from = Industry.Product, values_from = Value) 

industry_stock$Combined <- as.numeric(industry_stock$`B  Mining and quarrying`) + as.numeric(industry_stock$`M  Professional, scientific and technical activities`)

industry_stock$Combined <- as.character(industry_stock$Combined)

industry_stock <- industry_stock %>%
  pivot_longer(!Year, names_to = "Industry.Product", values_to = "Value")

industry_stock$Year <- gsub("X_", "", industry_stock$Year)
industry_stock$Year <- as.numeric(industry_stock$Year)
industry_stock$Value <- as.numeric(industry_stock$Value)

oilplotdata <- industry_stock %>%
  filter(Industry.Product == "Combined")
```

```{r, out.width = '65%', fig.align = 'left'}
oilplot <- ggplot(oilplotdata, aes(x = Year, y = Value)) + 
  geom_point() +
  geom_line() +
  ylim(0, 100) +
  labs(title = "Combined FDI from mining and R&D", 
       subtitle = "Data from wiiw",
       y = "Percent of FDI Stock")

oilplot
```

In other words, although the Russia-Ukraine conflicts and Western sanctions on Russia started in 2014, it is possible that the spike in 2014 may be from other sources, as it is difficult to move a lot of investment out that quickly, especially in sectors that require as much labor and infrastructure as the oil industry. 

We also could find less clear evidence that could still provide support for H1. For example, if we find that certain companies scale down their operations in Russia (or, for that matter, Ukraine) in 2013 or 2014 and then scale up their operations in Kazakhstan directly afterwards, this would appear to lend some credence to the "external stimuli" hypothesis, even if not directly stated. Additionally, a transfer of operations from a country at war to a country not at war is not the only way that war could affect MNC activity. Perhaps the presence of war in neighboring countries pushed US companies to invest in Kazakhstan for other reasons: maybe, for example, there was a movement among US MNCs to grow US involvement in Kazakhstan to *preserve* Kazakhstan as a space where US-based companies could operate without fear, following logic described by Johns and Wellhausen (2016).[^10] Or, perhaps, 

Moving to alternative hypotheses, how would I be convinced of H2's veracity? If a detailed analysis of Kazakhstani official documents and speeches shows that the Nazarbayev administration actively worked to encourage increased investment specifically from the United States around this time period, this would be fairly solid evidence for H2. The Nazarbayev administration pushing for increased western investment in general as well would support this hypothesis; maybe US companies followed behind other MNCs with relatively similar home environments and corporate philosophies. Looking at other countries' trends in FDI stock in Kazakhstan at the time, we can observe a rise in Dutch FDI stock as well, for example. Finally, evidence for H2 could take more indirect forms: perhaps the Nazarbayev administration did not say anything specific about attempting to attract more Western FDI; maybe some laws were just changed that encouraged more Western MNCs to invest, such as anti-corruption laws, lower regulations, etc. These sorts of evidence would help support H2.

There is potential evidence that would help me believe in H3's relevance as well. If we find that the US government actively encouraged certain actions by US-based MNCs and that US-based MNCs actually followed these directions, there would be strong support for H3. American firms are certainly not state-run, but it is possible that the government had a say, especially with regard to the sanctions regime against Russia that was coming out at the time. Additionally, if evidence shows that something specific about the US's approach toward either foreign direct investment as a whole or FDI specifically regarding Kazakhstan changed and US-based MNCs reacted in kind, this would support H3 as well. We know that the Barack Obama administration was more externally-focused than other US administrations; while some of the US FDI growth in Kazakhstan rest occurred during the Trump administration, it is possible that path dependency led firms to just continue their momentum from the Trump period. I consider the possibility of finding this sort of evidence unlikely given that Obama had been in office for several years by the time that the spike started; nonetheless, it is possible that Obama-era policies had some impact on firm activities in Kazakhstan.

### Conclusions

Overall, this paper will provide valuable contributions to a number of different areas of scientific and academic inquiry. Most importantly, it will utilize a single-country case study (Kazakhstan) to evaluate aspects of an understudied aspect of the IPE literature -- how war impacts foreign investment. By looking at Kazakhstan, a country adjacent to Russia, we can gain greater insights on the degree to which firms move from country to country due to war, and how that process looks. Additionally, using process tracing on a single-country case study, we can examine detailed evidence and causal-process mechanisms to better understand not just how firms deal with war in a given region, but also deal with war specifically in the post-Soviet context. 

Secondly, this study will shed light on influential internal processes occurring in both Kazakhstan and the United States during the middle of the last decade. For scholars and policymakers interested in learning more about politics in these two countries -- especially during a tumultuous period in parts of the former Soviet Union -- this paper will provide interesting information and useful insights that should help scholars and practitioners better deal with these sorts of questions in the future. Especially given the current conflict between Russia and Ukraine and the implementation of the "mother of all sanctions" on Russia, research like this should prove timely not just now, but in the future as well. 

Finally, the paper will address broader, more fully-studied aspects of the FDI literature and provide an intriguing data point for scholars to reference in future theoretical and practical work. The paper addresses questions such as a) why companies choose to engage in foreign direct investment in general, b) what internal factors within a host country impact the relative attraction of FDI, c) what internal factors within firms' host country impact their decisions to invest in one place or another, and d) how international events impact firms' calculus to increase or decrease their level of investment in a given location. Through all of these contributions, this paper will be a useful addition to the FDI literature, to the broader IPE literature, and to our general knowledge about Kazakhstan, the United States, and the post-Soviet region as a whole.[^11]

### Bibliography

\singlespacing

Beazer, Quintin H., and Daniel J. Blake. “The conditional nature of political risk: How home institutions influence the location of foreign direct investment.” *American Journal of Political Science* 62, no. 2 (2018): 470-485.

Bennett, Andrew and Alexander George. *Research Design Tasks in Case Study Method.* Paper presented at the MacArthur Foundation Workshop on Case Study Methods, Belfer Center for Science and International Affairs (BCSIA), Harvard University (October 17-19, 1997).

Bussmann, Margit. “Foreign Direct Investment and Militarized International Conflict.” *Journal of Peace Research* 47, no. 2 (2010): 143–53. 

Collier, David. “Process Tracing: Introduction and Exercises.” To Accompany Rethinking Social Inquiry, 2nd Edition. Beta Version (September 22, 2010).

Delios, Andrew, and Witold Jerzy Henisz. "Political Hazards, Experience, and Sequential Entry Strategies: The International Expansion of Japanese Firms, 1980–1998." *Strategic Management Journal* 24, no. 11 (2003): 1153-1164.

Dunning, John. Explaining International Production (London: Unwin Hyman, 1988).

Dunning, John. Multinational Enterprises and the Global Economy (Reading: Addison-Wesley, 1993), 144.

Jensen, Nathan M. “Democratic governance and multinational corporations: Political regimes and inflows of foreign direct investment.” *International Organization* 57.3 (2003): 587-616.

Jensen, Nathan M. *Nation-States and the Multinational Corporation: Political Economy of Foreign Direct Investment*. Princeton, NJ: Princeton University Press (2006).

Johns, Leslie, and Rachel L. Wellhausen. “Under One Roof: Supply Chains and the Protection of Foreign Investment.” *American Political Science Review* 110, no. 1 (2016): 31–51. 

Kerner, Andrew. “What we talk about when we talk about foreign direct investment.” *International Studies Quarterly* 58, no. 4 (2014): 804-815.

Kobrin, Stephen. “Testing the Bargaining Hypothesis in the Manufacturing Sector in Developing Countries,” *International Organization* 41, no. 4 (1987):
609–38.

Levy, Brian, and Pablo Spiller. "The Institutional Foundations of Regulatory Commitment: A Comparative Analysis of Telecommunications Regulation," *The Journal of Law, Economics, and Organization* 10, no. 2 (1994): 201-246.

Li, Quan, and Adam Resnick. “Reversal of Fortunes: Democratic Institutions and Foreign Direct Investment Inflows to Developing Countries.” *International Organization* 57, no. 1 (2003): 175–211. 

Li, Quan. “Foreign Direct Investment and Interstate Military Conflict.” *Journal of International Affairs* 62, no. 1 (2008): 53–66.

Li, Quan. “Democracy, autocracy, and expropriation of foreign direct investment.” *Comparative Political Studies* 42, no. 3 (2009): 1098-1127.

Lynch, Julia. “Can One Country Be Better Than Two for Comparative Politics?” *Italian Politics and Society* 60 (2005): 8-10.

Mankiw, N. Gregory et. al. "A Contribution to the Empirics of Economic Growth." *The Quarterly Journal of Economics* 107, no. 3 (1992): 407-437.

Markusen, James R. Multinational Firms and the Theory of International Trade (Cambridge: MIT Press, 2002).

Morisset, Jacques and Neda Pirnia. “How Tax Policy and Incentives Affect Foreign Direct Investment: A Review,” World Bank Working Paper, 2509, 2000; OECD, *Corporate Tax Incentives for Foreign Direct Investment,* no. 4, Paris: OECD (2001).

Nigh, Douglas. "The Effects of Political Events on United States Direct Foreign Investment: A Pooled Time-Series Cross-Sectional Analysis." *Journal of International Business Studies* 16, no. 1 (1985): 1-17.

Putnam, Robert D. “Diplomacy and Domestic Politics: The Logic of Two-Level Games.” *International Organization* 42, no. 3 (1988): 427–60. 

Staats, Joseph L. and Glen Biglaiser. "Foreign Direct Investment in Latin America: The Importance of Judicial Strength and Rule of Law." *International Studies Quarterly* 56: 193-202.

Walter, Andrew and Guathan Sen. *Analyzing the Global Political Economy* (Princeton: Princeton University Press, 2009)

```{r}

# ADDITIONAL PLOT!!

flowplot <- ggplot(fdiinflow, aes(x = Year, y = Value, color = Partner)) +
  geom_point() +
  geom_line() +
  labs(title = "Kazakhstan FDI Inflow, 2003-2021",
       subtitle = "Data from wiiw",
       y = "FDI Inflow in EUR m",
       x = "Year")
```


[^1]: John Dunning, Explaining International Production (London: Unwin Hyman, 1988).
[^2]: James R. Markusen, Multinational Firms and the Theory of International Trade (Cambridge: MIT Press, 2002).
[^3]: Data from UNCTAD.  
[^4]: See also Li, 2006; Li & Resnick, 2003. 
[^5]: Quan Li, “Foreign Direct Investment and Interstate Military Conflict." *Journal of International Affairs* 62, no. 1 (2008): 53–66.
[^6]: *Ibid*, 60. 
[^7]: David Collier, “Process Tracing: Introduction and Exercises.” To Accompany Rethinking Social Inquiry, 2nd Edition. Beta Version (September 22, 2010).
[^8]: RFE/RL, "More International Firms Relocate From Russia To Kazakhstan Over War In Ukraine." *RFE/RL*, December 12, 2022. https://www.rferl.org/a/kazakhstan-companies-relocate-russia-war/32173170.html.
[^9]: Due to changes in methodology, certain oil-related activities labeled as "geological exploration" were at one point placed under R&D; in 2016, wiiw relabeled them as "extraction of crude petroleum." To avoid ambiguity, I just combined the two categories for purposes of this study; this strategy more or less places upstream, downstream and R&D FDI together.
[^10]: In other words, by adding additional US MNC presence in Kazakhstan, companies become more tightly linked together, increasing the costs that would come with potential expropriation in the case of a future conflict involving Kazakhstan. At the time, there was some worry that Kazakhstan could also be a target for Russia, given its large ethnic Russian population in the north of the country.
[^11]: Note: for code, data, and some additional graphs, see my .rmd file at https://github.com/dfshapir/kazakhstan_ipe.